---
title: "viz3"
author: "Sarahy Martinez"
date: "2024-10-03"
output: github_document
---

```{r setup, include=FALSE}

library(tidyverse)

knitr:: opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal()+ theme(legend.position = "bottom"))

options(
  
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_fill_discrete = scale_fill_viridis_d()    

scale_colour_discrete = scale_fill_viridis_d() 

```


## Load the weather data

```{r}

weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") %>% 
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%   #extract the month variable, lubridate package tidyverse adjacent says to get the floor of a date aned sort of round whatever date or pull the dates donwn to the nearest. month is rounded down
  select(name, id, everything())


weather_df

```


## ` group_by`

```{r}
weather_df %>% 
  group_by(name) #appears like nothing happened by whats different says groups name[3]



weather_df %>% 
  group_by(name, month) %>% #groups by name and month
  ungroup() #gets rid of whatever grouping structure you have 



```


## counting the things

ie. how many ovbservations exist 
count months in this code 

```{r}

weather_df %>% 
  group_by(month) %>% 
  summarize(n_obs = n())   #n() counts for you, n_obs is the variable you created and end result is a tibble 



```



we can also use count()


```{r}

weather_df %>% 
  count(month, name = "n_obs")
```

**NEVER** use base R's `table` 


```{r, eval =FALSE}
  weather_df %>% 
  pull(month) %>% 
  table()  #painful to look at because the result isnt a dataframe, to get to the table function 
```



other helpful counters 

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    n_obs = n(),
    n_days = n_distinct(date) #n_distint looks at this month how many dated are distinct 
  )

```


## A digression on 2x2 tables 

```{r}
weather_df %>% 
  filter(name!= "Molokai_HI") %>% 
  mutate(
    cold = case_when(
      tmax< 5~ "cold",
      tmax >= 5 ~ "not_cold",  # ~ if this is true 
      TRUE ~ ""
    )
  ) %>% 
  group_by(name, cold) %>% 
  summarize( count = n())   #2x2 but not helpful, can also do a pivot wider but below is better
```

```{r}

weather_df %>%  
  mutate(cold = case_when(
    tmax <  5 ~ "cold",
    tmax >= 5 ~ "not_cold",
    TRUE     ~ ""
  )) %>% 
  filter(name != "Molokai_HI") %>%  
  janitor::tabyl(name, cold)            #this structure is better for 2x2
```



## General Summarries 

You can do lots of summaries

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE), #na.rm removes the missing variables
    mean_prcp = mean(prcp, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE)
     )

```


This is a dataframe! SO we can make a plot

```{r}

weather_df %>% 
  group_by(name,month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE), #na.rm removes the missing variables
    mean_prcp = mean(prcp, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE)
     ) %>% 
  ggplot(aes( x= month, y=mean_tmax, color = name))+ 
  geom_point() +
  geom_line()


```

Suppose you want to summarize many columns.

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(across(prcp: tmin, mean)) #can list a range and use across for each of these columns use this summary

```


Reminder: sometimes your results are easier to read in another format ,

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(mean_tmax = mean(tmax)) %>% 
  pivot_wider(
    names_from= name,
    values_from = mean_tmax
  )%>% 
# the names came from names, values came from tmax, and row name is whatever is left in your dataset (using pivot wider)
  knitr::kable(digits =1)  #formats r markdown and allows to print nicer

```

Now, another funcion affected by group_by is mutate 

## `group_by` and `mutate`

if you forget that you grouped can get into trouble

```{r}

weather_df %>% 
  group_by(name) %>% 
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE), #GROUPING when you do a mutate it will compute some observation and define a new variable according to this group first and then and this group etc. Why care because we might want to center to remove vertical distance.
 centered_tmax = tmax - mean_tmax  # we didn't subtract same mean from everyone just waikiki from waikiki etc. 
     ) %>% 
  ggplot(aes(x=date, y = centered_tmax, color=name))+
  geom_point()

```


about window functions 

so in the example above we took the mean and it took every t_max value observation and based on this tmax it then plugged in that for each row in our dataset.

In windown functions you put in  365 values and it spits out 365 values and these might be different values. 


ranking ...
```{r ranking}

weather_df %>% 
  group_by(name, month) %>%  #group first
  mutate(
    temp_rank = min_rank(tmax)) %>%   #we want to define a ranking according to the min_ rank tmax, temp rank we put the tmax values and says its the 22nd and 12 lowest values then the 5, 4th, 3 coldest days in the month. then filter for coldest day in each month. 
  filter(temp_rank == 1) # spits out rank 


#opposite order
weather_df %>% 
  group_by(name, month) %>%  #group first
  mutate(temp_rank = min_rank(desc(tmax)))
```


Can we take all of these oberservation, shift them and then drop them down.

```{r}
weather_df %>% 
  group_by(name) %>%  #group first
  mutate(lag_temp = lag(tmax,5))
#if you wanna compute changes from this day or this, or does yesterdays temp predict today 


weather_df %>% 
  group_by(name) %>%  
  mutate(temp_change = tmax - lag(tmax)) %>%  # we can also compute summaries
summarize(
  temp_change_max = max(temp_change, na.rm = TRUE),
   temp_change_sd = sd(temp_change, na.rm = TRUE)
)

```

Helpful because for example can you please compute the the day to day temperature variability, the day to day standard 
deviations and temperatures for each group. It has already been done above. We are answering a relatively complex question in a short amount of code. 











