---
title: "viz_i"
author: "Sarahy Martinez"
date: "2024-09-27"
output: github_document
---

```{r}
library(tidyverse)
library(ggridges)
library(hexbin)

```


# Load the Weather data 
We’ll be working with NOAA weather data, which is downloaded using rnoaa::meteo_pull_monitors function in the code chunk below; similar code underlies the weather dataset used elsewhere in the course. Because this process can take some time, I’ll cache the code chunk.
- create data frame not imported from reader instead rnoaa:: meteo_pull , the rnoa allows to access aggregated public service data, using it as a portal and telling I want data from these station and the min and max temps. then start date and final date. code goes online and says bring me this data. then name= to rename the weather stations. the tmin = tmin/10 bc temps from rnoa are 10th of degree celcuis and end with full degree max.then just relocate. 

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())


weather_df
```

As always, I start by looking at the data; below I’m showing the result of printing the dataframe in the console, but would also use view(weather_df) to look around a bit.

#Scatterplots! 


Create my first scatterplot ever.

```{r}

ggplot(weather_df, aes(x = tmin, y = tmax )) + 
  geom_point()

```


New approach, same plot.

```{r}

#good way because can also data manipulate 
weather_df %>% 
  ggplot(aes(x = tmin, y= tmax))+
  geom_point()

```

Save and edit plot object 

```{r}
weather_plot = 
  weather_df %>% 
  ggplot(aes(x= tmin, y = tmax))   #plot created and saved

weather_plot  #now its saved and we can change it later on, can also print it.
```

## Advanced Scatter Plot 

```{r}
  weather_df %>% 
  ggplot(aes(x= tmin, y = tmax, color= name))+ geom_point() #adding aes,

```

Adding new geometry, to give line that goes through, geom_smooth and add ste= false

```{r}
  weather_df %>% 
  ggplot(aes(x= tmin, y = tmax, color= name)) +
  geom_point()+
  geom_smooth(se = FALSE)
```

What about the aes placement? 
we specified inside teh ggplot but we can also do it in geom_point 

```{r}
  weather_df %>% 
  ggplot(aes(x= tmin, y = tmax)) +
  geom_point(aes(color= name))

# not much of a difference but it will start once we add geom_smooth 

```

```{r}

 weather_df %>% 
  ggplot(aes(x= tmin, y = tmax)) +
  geom_point(aes(color= name))
  geom_smooth()
  
  #color only applied to the scatter plot, we can define aes in ggplot and geom
  ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) +
  geom_smooth(se = FALSE)
```


Lets facet some things! 

```{r}
 weather_df %>% 
  ggplot(aes(x= tmin, y = tmax, color= name)) +
  geom_point()+
  geom_smooth(se = FALSE)
  facet_grid(. ~ name)  #we want a multipanel plot
  
  #same thing just not using pipe function 
  ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = .5) +   #adjust the alpha level for transperancy
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name) # . means don't create rows, instead columns that describe name vice versa
```


We can also see if we do alpha= tmin in the code 

```{r}
  ggplot(weather_df, aes(x = tmin, alpha = tmin, y = tmax, color = name)) + 
  geom_point() +  
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)


# tmin more transparent at the bottom and darker at the top 

#global option when do the # geom_point (alpha=0.2)


  ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha= 0.2) +  
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)


```


Let's combine some elements and try a new plot

```{r}

weather_df %>% 
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = prcp),alpha = 0.5) +  #shows the points in size perspective with aes(size= prcp)
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)

```

## Some small notes 

How many geoms have to exist?
```{r}
weather_df %>% 
  ggplot(aes( x = tmin, y= tmax, color = name)) +
  geom_smooth(se= FALSE)
#is a scatterplot required? no but a plot with just have smooth curve with no data points, but geom_point shows points. Here it just shows the smooth lines. 
```

You can use a neat geom !

```{r}

weather_df %>% 
  ggplot(aes(x = tmin, y= tmax)) +
  geom_hex()   #makes a hexagonal data with gradient, brighter color = lots of data points and the other not                 as much 
```

















