---
title: "Visualization"
subtitle: "Part II"
author: "Sarahy Martinez"
date: "2024-10-03"
output: github_document
---

## Look at the weather data 


```{r}
library(tidyverse)
library(patchwork)

```

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())


```

## Remember this plot?

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)  #3 aesthetics defined  tmin, tmax, color 

```

If this were a plot we wanted to share its sufficient but someone might not know whats going on so we will add a text or label.

## Labels 

```{r}

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)+
  labs(
    title = "Temperature Plot",
    x = "Minimum Daily Temperature (C)",
    y = "Maximum Daily Temperature (C)",
    caption = " Data from the rnoaa package; temperatures in 2017"
  )
```




## Scales 
It gives you a way to control what happens between x= tmin and y=tmax

```{r}

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)+
  labs(
    title = "Temperature Plot",
    x = "Minimum Daily Temperature (C)",
    y = "Maximum Daily Temperature (C)",
    caption = " Data from the rnoaa package; temperatures in 2017"
  )+
  scale_x_continuous(
    breaks=c(-15, 0 , 15),
    labels = c("-15c", "0", "15")   # need to be same length as breaks, more informative ticks 
  ) +     #categorical or factor use scale_x_discrete, defines breaks
scale_y_continuous(
  trans = "sqrt",   #transform data from linear scatter to sqrt, can also do log 
  position = "right"  # moves table name to the right
)

```



## Look at scale colors 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)+
  labs(
    title = "Temperature Plot",
    x = "Minimum Daily Temperature (C)",
    y = "Maximum Daily Temperature (C)",
    caption = " Data from the rnoaa package; temperatures in 2017"
  )+
  scale_color_hue(h = c(100,300))   #describes the feature of colors, control over colors 


```


```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)+
  labs(
    title = "Temperature Plot",
    x = "Minimum Daily Temperature (C)",
    y = "Maximum Daily Temperature (C)",
    caption = " Data from the rnoaa package; temperatures in 2017"
  )+
  scale_color_hue( name = "Location",
                   h =c(100,300)) #updated from variable "Name" to location 
 

```


scale_color_hue is meh, instead 

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)+
  labs(
    title = "Temperature Plot",
    x = "Minimum Daily Temperature (C)",
    y = "Maximum Daily Temperature (C)",
    caption = " Data from the rnoaa package; temperatures in 2017"
  )+
  viridis:: scale_color_viridis(   # scale has to be discrete and we have to tell it that
    
    name = "Location",
    discrete = TRUE
  )

```


Moving away from things that control specific mappings and not data dependent 

## Themes 

doesn't really depend on anything. Where the caption is absent on the data or AES mapping, deals with overall structure. 
```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)+
  labs(
    title = "Temperature Plot",
    x = "Minimum Daily Temperature (C)",
    y = "Maximum Daily Temperature (C)",
    caption = " Data from the rnoaa package; temperatures in 2017"
  )+
  viridis:: scale_color_viridis(   # scale has to be discrete and we have to tell it that
    
    name = "Location",
    discrete = TRUE )+
      theme(legend.position = "bottom")   #moved legend at the bottom , there are other arguments 


```


Suppose we didn't like the background etc. Changing the overall theme

```{r}
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)+
  labs(
    title = "Temperature Plot",
    x = "Minimum Daily Temperature (C)",
    y = "Maximum Daily Temperature (C)",
    caption = " Data from the rnoaa package; temperatures in 2017"
  )+
  viridis:: scale_color_viridis(   # scale has to be discrete and we have to tell it that
    
    name = "Location",
    discrete = TRUE )+
      theme_bw()  # made background white and bold outside , theme_classic no gridlines, ggthemes:: has many other themes

```

  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  