---
title: "viz_i"
author: "Sarahy Martinez"
date: "2024-09-27"
output: github_document
---

```{r}
library(tidyverse)
library(ggridges)

```


# Load the Weather data 
We’ll be working with NOAA weather data, which is downloaded using rnoaa::meteo_pull_monitors function in the code chunk below; similar code underlies the weather dataset used elsewhere in the course. Because this process can take some time, I’ll cache the code chunk.
- create data frame not imported from reader instead rnoaa:: meteo_pull , the rnoa allows to access aggregated public service data, using it as a portal and telling I want data from these station and the min and max temps. then start date and final date. code goes online and says bring me this data. then name= to rename the weather stations. the tmin = tmin/10 bc temps from rnoa are 10th of degree celcuis and end with full degree max.then just relocate. 

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())


weather_df
```

As always, I start by looking at the data; below I’m showing the result of printing the dataframe in the console, but would also use view(weather_df) to look around a bit.

#Scatterplots! 


Create my first scatterplot ever.

```{r}

ggplot(weather_df, aes(x = tmin, y = tmax ))




```


















